---
title: "Combinational Regularity Analysis with CORAtool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combinational Regularity Analysis with CORAtool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
library(CORAtool)
```

## The question this method asks

Most quantitative methods ask how much each variable contributes on average.
Combinational Regularity Analysis asks a different question: which
*combinations* of conditions are regularly followed by the outcome, allowing
that several different combinations may each be enough on their own.

Structures of that shape are called INUS structures: a condition that is
**I**nsufficient on its own but a **N**ecessary part of a combination that is
itself **U**nnecessary but **S**ufficient. Written out:

```
A{1}*B{0}*C{1} + D{1}*E{1}  =>  Y
```

Either "A is 1 and B is 0 and C is 1", or "D is 1 and E is 1", is enough for
`Y`. Neither is required. Asking what "the effect of A" is has no answer
here, because A only does anything alongside B and C.

CORA belongs to the family of configurational comparative methods, alongside
QCA and CNA. What distinguishes it is its starting point — switching circuit
analysis — and two capabilities that follow from it: multi-value conditions,
and *complex effects*, where several outcome columns are minimised jointly
rather than one at a time.

## The data

A data frame, one row per case. Conditions must be non-negative integers
**coded from zero upwards with no gaps**: `0, 1` for a binary condition,
`0, 1, 2` for a three-valued one. `cora_recode()` will do that for you, and
an analysis of data coded otherwise is refused rather than attempted — see
*Coding* below for why that matters.

```{r}
df <- data.frame(A   = c(1, 0, 1, 0),
                 B   = c(1, 0, 0, 1),
                 C   = c(0, 1, 1, 0),
                 OUT = c(1, 1, 0, 1))
ctx <- cora_context(df, output_labels = "OUT")
```

`cora_context()` holds the data together with every analytical choice. It
computes nothing on its own: the truth table, the prime implicants and the
solutions are each derived on first request and kept afterwards.

## Five stages

```{r}
cora_truth_table(ctx)
```

Cases sharing a combination of condition values are aggregated into one
configuration. `n_cut` sets how many cases a configuration needs before it is
believed, and `inc_score1` how consistently it must show the outcome to count
as positive.

```{r}
cora_prime_implicants(ctx)
```

Boolean minimisation reduces the positive configurations to prime
implicants — combinations that cannot be shortened without covering a
negative case. A `#` marks an **essential** prime implicant: one that holds a
positive case no other prime implicant reaches, so every solution must
contain it.

```{r}
cora_pi_chart(ctx)
```

The chart says which prime implicant covers which positive row. Petrick's
method then reads every irredundant solution off it — every combination that
covers all positive rows and stops covering them all if any term is removed.

```{r}
cora_irredundant_sums(ctx)
```

## Reading the result

Two solutions, not one. This is **model ambiguity**, and it is a property of
the data rather than a failure of the analysis: on every configuration that
was actually observed, these two make identical predictions. They differ only
on configurations nobody observed.

The honest response is to report both. Narrowing to one requires an argument
from outside the data — theory, timing, a design that rules a combination
out — because the data has already been used up.

```{r}
cora_pi_details(ctx)
```

| column | meaning |
|---|---|
| `Cov.r` | of all cases showing the outcome, the share this prime implicant covers |
| `Inc.` | of the cases this prime implicant covers, the share showing the outcome |
| `M1`, `M2` | within that solution, the share of positive cases **only** this term covers |
| `NA` | this prime implicant is not in that solution |

High inclusion means the combination is close to sufficient: when it holds,
the outcome nearly always follows. High coverage means it explains much of
what happened. They answer different questions, and a term with inclusion 1
and coverage 0.05 is telling you it never misfires and almost never fires.

A unique coverage of 0 is worth pausing on: that term covers nothing the
others do not already cover. It is in the solution because removing it would
break irredundance, not because it carries any case of its own.

```{r}
cora_system_details(ctx)
cora_describe(cora_irredundant_sums(ctx)[[1]])
```

`cora_describe()` turns the scores into a relation: `=>` for sufficient, `<=`
for necessary, `<=>` for both.

## Multi-value conditions

Nothing changes except the range of the values.

```{r}
tort <- cora_context(gross_carvin, "TORT", case_col = "Case",
                     algorithm = "ON-OFF")
cora_irredundant_sums(tort)
```

Every literal carries its value explicitly, binary conditions included:
`LENG{2}` is "LENG equals 2", `PRIC{0}` is "PRIC equals 0". Literals inside a
conjunction are written in alphabetical order of the condition, so the same
analysis prints the same string however the columns were arranged.

## Complex effects

Several outcomes can be minimised jointly. The result is an irredundant
*system*: one function per outcome, irredundant as a whole even where an
individual function is not.

```{r}
minaret <- cora_context(swiss_minaret, c("X", "M"), algorithm = "ON-OFF")
cora_irredundant_systems(minaret)[[1]]
```

`L{0}*T{0}` and `S{1}` serve both outcomes; `T{1}` serves only `M`. Which
mechanisms are shared and which are particular to one outcome is the question
complex effects exist to answer, and analysing each outcome separately cannot
reach it.

## Diagrams

```{r, fig.alt = "Two-level logic diagram of the tort liability solution"}
cora_logigram(cora_irredundant_sums(tort)[[1]])
```

Read it left to right. The vertical lines are condition buses, one per
condition that appears in the solution. A dot on a bus is a literal, labelled
with the value it takes. The yellow blocks are AND gates forming the
conjunctions; a term of a single literal has nothing to combine and runs
straight past them. The blue shield is the OR gate, and the name on its right
is the outcome.

The expression and its scores are printed above the drawing, so the figure
and the formula travel together. `show_terms = TRUE` labels each gate with
the conjunction it forms; `title` and `subtitle` take your own text, or `NA`
for none.

## Choosing the algorithm

`"ON-DC"` is the classical Quine-McCluskey algorithm over positive and
don't-care terms; `"ON-OFF"` is McCluskey's modified algorithm over positive
and negative terms. On data coded from zero they return the same prime
implicants.

They do not cost the same. `"ON-DC"` expands the full configuration space, so
its cost grows exponentially in the number of levels of a single condition:
roughly a second at twelve levels, half a minute at eighteen, out of reach at
thirty. `"ON-OFF"` works from the observed rows and answers in a fraction of
a second at any size. **Prefer `"ON-OFF"` when a condition has many levels or
there are many conditions.** A condition of more than twelve levels draws a
warning saying so.

## Coding

Conditions must run `0, 1, 2, ...` because the algorithms take the number of
distinct values as the range of values. A condition coded `{1, 2}` has two
distinct values, so an implicant that does not mention it is filled in with
`{0, 1}` — a set containing a value never observed and missing one that was.
Cases are then dropped from coverage for no reason connected to the
implicant.

```{r, error = TRUE}
gapped <- data.frame(A = c(1, 1, 0, 0), B = c(2, 1, 2, 2),
                     OUT = c(1, 1, 0, 1))
cora_prime_implicants(cora_context(gapped, "OUT"))
```

```{r}
fixed <- cora_recode(gapped, "B")
fixed
```

Recoding shifts the labels: `B{2}` becomes `B{1}`. The structure is
unchanged, but a value in a write-up has to be read against the coding that
produced it.

## When there are too many solutions

A chart with a few dozen prime implicants can have tens of thousands of
irredundant solutions. Every one is valid, which is exactly the problem:
nobody can report them all. It usually means too many conditions or too loose
a threshold.

`max_depth` asks a narrower question, and asks it of the search rather than
of the finished list, which is often the difference between an answer and no
answer:

```{r}
praet <- cora_context(bergschlosser, "PRAET",
                      input_labels = c("AGRPOP", "PARCL", "APROG",
                                       "PS", "RQ", "LRC"),
                      case_col = "Case", inc_score1 = 0.6,
                      algorithm = "ON-OFF")
length(cora_irredundant_sums(praet, max_depth = 7))
```

Unrestricted, that analysis has 74,524 solutions and takes about half a
minute; restricted to seven prime implicants it answers in a fraction of a
second. The restriction applies to the call, not to the context, so a later
unrestricted call still sees everything.

## Searching for a smaller model

`cora_data_mining()` runs the analysis over every combination of a given
number of conditions, which is a configurational Occam's razor: how few
conditions still generate a solution?

```{r}
cora_data_mining(mccluskey, c("F1", "F2"), len_of_tuple = 2)
```

`automatic = TRUE` widens the search one condition at a time until a
combination yields a solution.

## Relation to the Python implementation

This package is an R port of the Python packages CORA and LOGIGRAM. It
computes in plain R and needs no Python installation. Prime implicants,
coverage sets and solution sets agree with the original across its own test
data and 200 randomly generated data sets.

It departs from the original in seven documented places. The departures that
change numbers are: data not coded from zero is refused rather than analysed;
an implicant's inclusion score is measured against the outcome it is an
implicant of, so the two algorithms agree; and a tuple whose only solution is
the tautology scores zero in data mining rather than a perfect one. Appendix A
of the extended manual records each with its source location and a
reproducible example, and the evidence behind the comparison is in
`tools/VALIDATION.md` in the source repository.

If the Python package is installed, `cora_compare_python()` will run both and
compare them.

## Caveats worth carrying

- CORA finds **regularity structure in data**. Moving from there to a causal
  claim needs what the method cannot supply: theory, timing, a design that
  rules out common causes.
- Limited diversity is the normal condition, not an edge case. With `k`
  conditions there are at least `2^k` possible configurations and you will
  have observed a small fraction. How much of a solution rests on
  configurations nobody saw is yours to know and to say.
- Coverage and inclusion are not p-values. There is no null hypothesis here
  and no significance to report.
- Report every solution, the algorithm, the thresholds, and what the values
  of each condition mean. Without the last of these a reader cannot tell what
  `LENG{2}` is a claim about.

## Extended manual

A fuller manual covers the theory, the output, the diagrams and the
comparison with the Python implementation, QCA, QCApro and cna in more depth
than this vignette. It ships in English and in Traditional Chinese:

```r
file.show(system.file("docs", "manual_en.md", package = "CORAtool"))
file.show(system.file("docs", "manual_zh-TW.md", package = "CORAtool"))
```

`inst/examples/getting-started.R` is the same ground as a runnable script.
